---
title: "Data 606 Final Project: New York City Road Condition Complaints and Demographics"
author: "Armenoush Aslanian-Persico"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
    theme: spacelab
  
---

# Introduction

## Topic

What is your research question? Why do you care? Why should others care?

Explanation of 311

Explanation of Census

## Cases
What are the cases, and how many are there?

Cases by zip code, combined data for 
182 zip codes in NYC
325025 street condition complaints with reported zip code in date range 

## Data Collection
Describe the method of data collection.

## Study Type
What type of study is this (observational/experiment)?
Explain how you've arrived at your conclusion using information on the sampling and/or experimental design.

## Sources

311 Complaint Data
NYC Open Data
311 Service Requests from 2010 to Present
https://nycopendata.socrata.com/
Selected 311 complaints from January 1, 2013 through May 12, 2016.
Filtered by Complaint category - "Street Condition" complaint only

Income Data
US Census
2010-2014 American Community Survey 5-Year Estimates
Selected economic characteristics
http://factfinder.census.gov/

Population and Demographic Data
US Census
2010-2014 American Community Survey 5-Year Estimates
ACS DEMOGRAPHIC AND HOUSING ESTIMATES
http://factfinder.census.gov/faces/tableservices/jsf/pages/productview.xhtml?pid=ACS_14_5YR_DP05&src=pt

---

# Data 

## Libraries
```{r, libraries, eval=TRUE, warning=FALSE, message=FALSE}
library(DT)
library(ggplot2)
library(knitr)
library(plotly)
library(plyr)
library(dplyr)
library(RCurl)
library(rmarkdown)
library(stringr)
library(tidyr)

```

```{r, libraries, eval=FALSE, warning=FALSE, message=FALSE}
library(XML)
library(scales)
library(RColorBrewer)
library(leaflet)
library(httr)
library(jsonlite)
```
## Response 
What is the response variable, and what type is it (numerical/categorical)?

## Explanatory
What is the explanatory variable(s), and what type is it (numerical/categorival)?

## Relevant summary statistics

182 zip codes in NYC
96710 mean income by zip code for NYC
325025 street condition complaints with reported zip code in date range 

## Generalizability

Scope of inference
Identify the population of interest, and whether the findings from this analysis can be generalized to that population, or, if not, a subsection of that population. Explain why or why not. Also discuss any potential sources of bias that might prevent generalizability.

## Causality

Scope of inference
Can these data be used to establish causal links between the variables of interest? Explain why or why not.

---

# Analysis

## Transformation

### 311 Data

Selected 311 complaints from January 1, 2013 through May 12, 2016.
Millions of complaints across all city agencies: Health, Environmental Protection, Sanitation, Police, others.
Filtered to only Department of Transportation data.
Filtered to only complaints which had a zip code as part of the complaint.
Downloaded dataset was already all Street Condition complaints. This was further filtered to the 5 primarily asphalt-related street conditions and 2 markings-related street conditions.

Asphalt conditions:
Pothole	
Rough, Pitted or Cracked Roads	
Hummock	
Cave-in	
Failed Street Repair	

Marking conditions:
Line/Marking - After Repaving
Line/Marking - Faded

To further simplify, all asphalt conditions were combined into one roadway surface complaint group, named "asphalt."

Filled in missing values as 0.

### Zip Code and Borough Data

A list of zip codes and their associated borough was extracted from the 311 data.
Created table of zip codes and associated borough

### Income by Zip Code Data

Census Data

Downloaded New York State zip codes only.

Selected columns using metadata.

Converted zip column to display zip code only.

Added "ZIP_" prefix to prevent forcing to numeric value.

HC01_VC90 - Estimate; INCOME AND BENEFITS (IN 2014 INFLATION-ADJUSTED DOLLARS) - With earnings - Mean earnings (dollars)

### Population and Demographics by Zip Code Data

Census Data

Downloaded New York State zip codes only.

Selected columns using metadata.

Converted zip column to display zip code only.

Added "ZIP_" prefix to prevent forcing to numeric value.

Total population per zip code

HC01_VC03 - Estimate; SEX AND AGE - Total population
HC03_VC49 - Percent; RACE - One race - White
HC03_VC50 - Percent; RACE - One race - Black or African American

## Statistics

Perform relevant descriptive statistics, including summary statistics and visualization of the data. Also address what the exploratory data analysis suggests about your research question.

### Load Data

```{r loaddata, eval=TRUE}
sumzipdata <- getURL("https://raw.githubusercontent.com/spsstudent15/2016-01-606-Project/master/zipdatacsv2.csv")

sdata<- data.frame(read.csv(text=sumzipdata))
```

### Statistics: Summary

```{r sum1, eval=TRUE}
# Structure
# Summary
# Distribution of Road Condition Complaints Per Capita

str(sdata)
summary(sdata)

hist(sdata$pcomp, col="#fff9c4", breaks=10, main="Distribution of Road Condition Complaints Per Capita", xlab="Complaints Per Capita")
```

## Statistics: Income

```{r income, eval=TRUE}
# Box Plot of Income and Population by Zip Code
# Distribution of Income
# Road Condition Complaints vs. Income
# Road Condition Complaints vs. Income, with xlim1
# Road Condition Complaints vs. Income, with xlim2

boxplot(sdata$income, sdata$pop, names=c("Income","Population"), col=c("#c5e1a5","#ffe0b2"), main="Box Plot of Income and Population by Zip Code")

hist(sdata$income, col="#c5e1a5", breaks=20, main="Distribution of Income", xlab="Income")

minc<- lm(pcomp ~ income, data = sdata)
plot(sdata$pcomp ~ sdata$income, col="#81d4fa", main="Road Condition Complaints vs. Income", xlab="Income", ylab="Complaints Per Capita")
abline(minc)

minc1<- lm(pcomp ~ income, data = sdata)
plot(sdata$pcomp ~ sdata$income, col="#1976d2", main="Road Condition Complaints vs. Income, $20,000-$200,000", xlab="Income", ylab="Complaints Per Capita", xlim=c(20000,200000))
abline(minc1)

minc2<- lm(pcomp ~ income, data = sdata)
plot(sdata$pcomp ~ sdata$income, col="#01579b", main="Road Condition Complaints vs. Income, $30,000-$100,000", xlab="Income", ylab="Complaints Per Capita", xlim=c(30000,100000), ylim=c(0,0.10))
abline(minc2)

```

## Statistics: Race

```{r race, eval=TRUE}
# Correlation for Percent White
# Correlation for Percent Black
# Distribution of Percent Whites in Zip Code
# Distribution of Percent Blacks in Zip Code
# Road Condition Complaints vs. Percentage Whites
# Road Condition Complaints vs. Percentage Blacks
# Residuals for Percent White
# Residuals for Percent Black


wcor<-cor(sdata$pcomp, sdata$pwhite)
bcor<-cor(sdata$pcomp, sdata$pblack)

hist(sdata$pwhite, col="#ffcc80", breaks=20, main="White Population", xlab="Distribution of Percent Whites in Zip Code", ylim=c(0,70))

hist(sdata$pblack, col="#26a69a", breaks=20, main="Black Population", xlab="Distribution of Percent Blacks in Zip Code", ylim=c(0,70))

mwhite<- lm(pcomp ~ pwhite, data = sdata)
plot(sdata$pcomp ~ sdata$pwhite, col="#ffcc80", main="Road Condition Complaints vs. Percentage Whites", xlab="Percent White Population per Zip Code", ylab="Complaints Per Capita")
abline(mwhite)

mblack<- lm(pcomp ~ pblack, data = sdata)
plot(sdata$pcomp ~ sdata$pblack, col="#26a69a", main="Road Condition Complaints vs. Percentage Blacks", xlab="Percent Black Population per Zip Code", ylab="Complaints Per Capita")
abline(mblack)

plot(mwhite$residuals ~ sdata$pcomp, main="Residuals for Percent White", col="#ffcc80")
abline(mwhite)

hist(mwhite$residuals, main="Residuals for Percent White", ylim=c(0,80), xlab="Residuals" , col="#ffcc80")

plot(mblack$residuals ~ sdata$pcomp, main="Residuals for Percent Black", col="#26a69a")
abline(mblack)

hist(mblack$residuals, main="Residuals for Percent Black", ylim=(c(0,80)), xlab="Residuals", col="#26a69a")

```

The correlation for whites and per capita roadway complaints is `r round(wcor,2)`.

The correlation for blacks and per capita roadway complaints is `r round(bcor,2)`.

## Statistics: Borough

```{r boro, eval=TRUE}
# Box Plots by Borough

# Per Capita Complaints by Borough
# Per Capita Complaints by Borough, Asphalt Only
# Per Capita Complaints by Borough, Missing Markings Only
# Per Capita Complaints by Borough, Faded Markings Only


boxplot(
    sdata$pcomp[sdata$borough=="BRONX"],
    sdata$pcomp[sdata$borough=="BROOKLYN"],
    sdata$pcomp[sdata$borough=="MANHATTAN"],
    sdata$pcomp[sdata$borough=="QUEENS"],
    sdata$pcomp[sdata$borough=="STATEN ISLAND"],
    col=(c("#E8EAF6","#c5cae9","#9fa8da","#7986cb","#5c6bc0")),
    names=(c("BRONX","BKLYN","MANHATTAN","QUEENS","STATEN IS.")), 
    main="Per Capita Complaints by Borough", ylab="Per Capita Complaints", xlab="Borough")


boxplot(
    sdata$pcompa[sdata$borough=="BRONX"],
    sdata$pcompa[sdata$borough=="BROOKLYN"],
    sdata$pcompa[sdata$borough=="MANHATTAN"],
    sdata$pcompa[sdata$borough=="QUEENS"],
    sdata$pcompa[sdata$borough=="STATEN ISLAND"],
    col=(c("#E8EAF6","#c5cae9","#9fa8da","#7986cb","#5c6bc0")),
    names=(c("BRONX","BKLYN","MANHATTAN","QUEENS","STATEN IS.")), 
    main="Per Capita Complaints by Borough, Asphalt Only", ylab="Per Capita Asphalt Complaints", xlab="Borough")

boxplot(
    sdata$pcompm[sdata$borough=="BRONX"],
    sdata$pcompm[sdata$borough=="BROOKLYN"],
    sdata$pcompm[sdata$borough=="MANHATTAN"],
    sdata$pcompm[sdata$borough=="QUEENS"],
    sdata$pcompm[sdata$borough=="STATEN ISLAND"],
    col=(c("#E8EAF6","#c5cae9","#9fa8da","#7986cb","#5c6bc0")),
    names=(c("BRONX","BKLYN","MANHATTAN","QUEENS","STATEN IS.")), 
    main="Per Capita Complaints by Borough, Missing Markings Only", ylab="Per Capita Missing Markings Complaints", xlab="Borough")

boxplot(
    sdata$pcompf[sdata$borough=="BRONX"],
    sdata$pcompf[sdata$borough=="BROOKLYN"],
    sdata$pcompf[sdata$borough=="MANHATTAN"],
    sdata$pcompf[sdata$borough=="QUEENS"],
    sdata$pcompf[sdata$borough=="STATEN ISLAND"],
    col=(c("#E8EAF6","#c5cae9","#9fa8da","#7986cb","#5c6bc0")),
    names=(c("BRONX","BKLYN","MANHATTAN","QUEENS","STATEN IS.")), 
    main="Per Capita Complaints by Borough, Faded Markings Only", ylab="Per Capita Faded Markings Complaints", xlab="Borough", ylim=c(0,0.006))
```


Comparing the mean of the distributions by splitting the variable into the borough groups.

```{r boromeans, eval=TRUE}
# means by borough

by(sdata$pcomp, sdata$borough, mean)
by(sdata$pcomp, sdata$borough, summary)
```


## Outliers

High income

ZIP_10803 232513	BRONX actually Westchester county

ZIP_11109 168940	QUEENS Long Island City new skyscraper condos on the waterfront

Highest complaints per capita
ZIP_10004 Lower Manhattan, includes the NYC DOT headquarters

ZIP_11363 Little Neck, Queens


Note on missing/incorrect zip codes showing as 0 income or NA income


## Inference
If your data fails some conditions and you can't use a theoretical method, then you should use simulation. If you can use both methods, then you should use both methods. It is your responsibility to figure out the appropriate methodology.

Are the conditions for inference satisfied?

Sample size n>30
Unimodal normal distribution
Random sample

> *Check if inference conditions are satisfied.*

```{r inf1, eval=FALSE}
#categorical
by(nc$weight, nc$habit, length)
hist(nc$weight, col="green")
```


Write the hypotheses for testing if the average weights of babies born to smoking and non-smoking mothers are different.

> *The null testing hypothesis would be that there is no difference between the average weights of babies born to smokers and nonsmoker mothers. The alternate hypothesis would be that there is a difference in the average weights between these two groups.*

Next, we introduce a new function, `inference`, that we will use for conducting
hypothesis tests and constructing confidence intervals. 

```{r inf2, eval=FALSE}
#categorical
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical")
```

 
The first argument is `y`, which is the response variable that we are 
interested in: `nc$weight`. The second argument is the explanatory variable, 
`x`, which is the variable that splits the data into two groups, smokers and 
non-smokers: `nc$habit`. The third argument, `est`, is the parameter we're 
interested in: `"mean"` (other options are `"median"`, or `"proportion"`.) Next
we decide on the `type` of inference we want: a hypothesis test (`"ht"`) or a 
confidence interval (`"ci"`). When performing a hypothesis test, we also need 
to supply the `null` value, which in this case is `0`, since the null 
hypothesis sets the two population means equal to each other. The `alternative`
hypothesis can be `"less"`, `"greater"`, or `"twosided"`. Lastly, the `method` 
of inference can be `"theoretical"` or `"simulation"` based.


Change the `type` argument to `"ci"` to construct and record a confidence 
interval for the difference between the weights of babies born to smoking and 
non-smoking mothers.

By default the function reports an interval for ($\mu_{nonsmoker} - \mu_{smoker}$)
. We can easily change this order by using the `order` argument:

```{r inf3, eval=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical", 
          order = c("smoker","nonsmoker"))
```


## Check conditions

Theoretical inference (if possible) - hypothesis test and confidence interval
Simulation based inference - hypothesis test and confidence interval
Brief description of methodology that reflects your conceptual understanding

### Hypothesis test

### Confidence interval

---

# Conclusion: 

## Summary
Write a brief summary of your findings without repeating your statements from earlier. 

## Insights
Also include a discussion of what you have learned about your research question and the data you collected. 

Self-selection of 311 reporters
Why people may  not want to call 311
Some of these reports are not available on the mobile app and require the web interface or a phone call to report.

Median income would have been better than mean since high incomes in large cities skews the average higher. In reality, close to 50% of NYC is living below or near the poverty line, a very different story from a mean income of $90,000.

Missing and faded markings might mean the same thing to 311 callers. For operational purposes I was looking for accurate data since these are handled through different processes.

## Future Research
You may also want to include ideas for possible future research.